Self Referencing Sequences

### 2014-11-07 draft.tex
density bounds for c = 12 finished

As discussed in the Section~\ref{substitutionrules}, it is unclear how to establish a set of rules for $C = (1 2)$. However, there is some structure that we can exploit. In particular, we know that the sequences $1, 1, 1$ and $2, 2, 2$ will never appear in the self-referential sequence. This is because such sequences would be part of a block of size at least $3$. There would either be a 3 earlier in the sequence, or $1,1,1$ or $2,2,2$ must be the first block. We know that neither option is possible.

What does this mean? Well, we know that if there is a limiting density of $1$'s and $2$'s, then that density $d_2$ is bounded between 
\[ \dfrac{1}{3} \leq d_2 \leq \dfrac{2}{3} \]

So now we know that there must be at least one $2$ for every three numbers. We can ask  the following similar question: how many $2$'s are in any consecutive substring of length $9$? We know there must be at least three $2$'s, but do we need more? There are only three cases where there are three 2's in a segment of $9$ numbers with every consecutive subsequence of length 3 containing at least one 2, namely
\[ 2, 1, 1, 2, 1, 1 ,2 , 1, 1 \]
\[ 1, 2, 1, 1, 2, 1, 1, 2, 1 \]
\[ 1, 1, 2, 1, 1, 2, 1, 1, 2 \]
In each of these cases, we can look at the block lengths. Since the sequence is self-referential, the consecutive subsequence of block lengths must have appeared earlier in the sequence. The block lengths for the three above sequences are:
\[ 2, 1, 2, 1, 2 \]
\[ 1, 2, 1, 2, 1\]
\[ 2, 1, 2, 1, 2 \] 
We can look at the block lengths again. If the above segments were in the self-referencing sequence, then so is $1, 1, 1$, a contradiction. This means that none of the above consecutive subsequences of length 9 are possible. This means that in every consecutive subsequence of length $9$, there must be at least $4$ 2's. In other words, in addition to having one 2 for every group of 3, there is another 2 for every group of 9. Therefore, the density must be bounded by
\[ \dfrac{1}{3} + \dfrac{1}{9} \leq d_2 \leq 1 - \dfrac{1}{3} - \dfrac{1}{9} \]
In general, we can continue this process for higher powers of $3$. We did it once more for consecutive subsequences of length 27 and found you needed thirteen 2's. Thus, the density bounded by
\[ \dfrac{1}{3} + \dfrac{1}{9} + \dfrac{1}{27} \leq d_2 \leq 1 - \dfrac{1}{3} - \dfrac{1}{9} - \dfrac{1}{27} \]
The number of cases to check increases very quickly: there are $3$ choices to put the $2$ in the groups of $3$, $3$ choices to put the 3 groups of 2's and 2 positions within these groups for the groups of 9, and so on. Its not hard to see that this number grows exponentially fast, making it computationally difficult to bound the density.

If there is a pattern which we can exploit with an inductive proof, we will be able to show that the density exists and is $\frac{1}{2}$. The density will be bounded below by $\sum_{n=1}^\infty \dfrac{1}{3^n} = \frac{1}{2}$ and bounded above by the same number. However, finding patterns in this procedure to improve the bound seems difficult.

>>>>

As discussed in the Section~\ref{substitutionrules}, it is unclear how to establish a set of rules for $C = (1 2)$. However, there is some structure that we can exploit. In particular, we know that the sequences $1, 1, 1$ and $2, 2, 2$ will never appear in the self-referential sequence. This is because such sequences either imply the existence of a 3 earlier in the sequence, or make up the first block of the sequence. Neither option is possible.

This simple observation implies that if there is a limiting density of $1$'s and $2$'s, then that density $d_2$ is bounded between 
\[ \dfrac{1}{3} \leq d_2 \leq \dfrac{2}{3} \]

So we know there must be at least one $2$ for every three consecutive terms of the sequence. We can ask  the following similar question: how many $2$'s are in any consecutive substring of length $9$? The answer is (trivially) at least 3, by our original observation. However, there are only three cases where there are exactly three 2's in a segment of $9$ numbers. In particular, they are
\[ 2, 1, 1, 2, 1, 1 ,2 , 1, 1 \]
\[ 1, 2, 1, 1, 2, 1, 1, 2, 1 \]
\[ 1, 1, 2, 1, 1, 2, 1, 1, 2 .\]
Note that we rule out all strings of 9 numbers with a consecutive run of three 1's or three 2's.

In each of the above cases, we can look at the block lengths. Since the sequence is self-referential, the consecutive subsequence of block lengths must have appeared earlier in the sequence. The block lengths for the three above sequences are:
\[ 2, 1, 2, 1, 2 \]
\[ 1, 2, 1, 2, 1\]
\[ 2, 1, 2, 1, 2 .\] 
We repeat this reasoning. If the above segments were in the self-referencing sequence, then so is $1, 1, 1$, a contradiction. This means that none of the above consecutive subsequences of length 9 are possible, which implies that in every consecutive subsequence of length $9$, there must be at least four 2's. In other words, in addition to having one 2 for every group of 3, there is another 2 for every group of 9. Therefore, the density must be bounded by
\[ \dfrac{1}{3} + \dfrac{1}{9} \leq d_2 \leq 1 - \dfrac{1}{3} - \dfrac{1}{9} \]
We can try to continue this process for higher powers of $3$. For subsequences of length 27, our result for subsequences of length 9 tells us that we must have at least twelve 2's. But we were able to show that a contradictino is reached in each of these cases, similar to the way a contradiction is reached above. So for consecutive subsequences of length 27, there must be at least thirteen 2's. Thus, the density bounded by
\[ \dfrac{1}{3} + \dfrac{1}{9} + \dfrac{1}{27} \leq d_2 \leq 1 - \dfrac{1}{3} - \dfrac{1}{9} - \dfrac{1}{27} \]
Unfortunately, we have not been able to generalize this reasoning, as the number of cases that must be checked increases exponentially fast. There are $3$ choices for where to put the $2$ in the groups of $3$, $3$ choices for where to put the 3 groups of 2's and 2 positions within these groups for the groups of 9, and so on.

If there exists a pattern that we can make general statements about for arbitrary powers of 3, then this will imply that the density is bounded below and above by
\[ \sum_{n=1}^\infty \dfrac{1}{3^n} = \frac{1}{2} \]
Note that this would also imply the existence of a density. The previous claims can only hope to bound the density, if it exists. The issue is that the ``limiting" density may not exist as it may oscillate between the two bounds. But if the bound is tight (upper and lower bounded at $1/2$) then there is no room for oscillation, and so the density would exist.
